A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-04-28 submission/manuscript.rmd
proof reading submission

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\vspace{20mm}

@@

\vspace{30mm}
**Funding**: This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 806968. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

\newpage
\newpage
# Abstract (247 words)
\singlespacing 
**Aim:** One of the aims of the Observation Health Data Sciences
and Informatics (OHDSI) initiative is population-level treatment effect
estimation in large observational databases. Since treatment effects are
well-known to vary across groups of patients with different baseline risk, we
aimed to extend the OHDSI methods library with a framework for risk-based
assessment of treatment effect heterogeneity.

**Materials and Methods:** The proposed framework consists of five steps: 1)
definition of the problem, i.e. the population, the treatment, the comparator
and the outcome(s) of interest; 2) identification of relevant databases; 3)
development of a prediction model for the outcome(s) of interest; 4) estimation
of propensity scores within strata of predicted risk and estimation of relative
and absolute treatment effect within strata of predicted risk; 5) evaluation and
presentation of results.

**Results:** We demonstrate our framework by evaluating heterogeneity of the
effect of angiotensin-converting enzyme (ACE) inhibitors versus beta blockers on
a set of 9 outcomes of interest across three observational databases. With
increasing risk of acute myocardial infarction we observed increasing absolute
benefits, i.e. from -0.03% to 0.54% in the lowest to highest risk groups.
Cough-related absolute harms decreased from 4.1% to 2.6%.

**Conclusions:** The proposed framework may be useful for the evaluation of
heterogeneity of treatment effect on observational data that are mapped to the
OMOP Common Data Model. The proof of concept study demonstrates its feasibility
in large observational data. Further insights may arise by application to safety
and effectiveness questions across the global data network.

\newpage 
\doublespacing 
\linenumbers

# Introduction
Interest in understanding how a treatment’s effect varies across patients—a
concept described as heterogeneity of treatment effects (HTE)—has been growing.
This concept is central to the agenda for both personalized (or precision)
medicine and comparative effectiveness research. More formally, HTE has been
defined as non-random variability in the direction or magnitude of a treatment
effect, in which the effect is measured using clinical outcomes
[@KRAVITZ2004]. Usually, analyses focus on the relative scale, where treatment
effects are assessed one at a time in patient subgroups defined from single
covariates, an approach that suffers from low power and multiplicity issues
[@Varadhan2013]. However, even with well-established constant relative
effects, treatment benefit (or harm) may vary substantially on the absolute
scale.

@@

The Observational Health Data Sciences and Informatics (OHDSI) collaborative
[@hripcsak2015observational] has established an international network of data
partners and researchers that aim to bring out the value of health data through
large-scale analytics by mapping all available databases to the Observational
Medical Outcomes Partnership (OMOP) Common Data Model (CDM) [@Overhage2012].
The common data structure enables analyses at a very large scale. For example,
in a recent study [@Suchard2019], a large set of first-line treatments for
hypertension was compared with respect to 55 outcomes in a network of databases,
including 4.9 million patients from around the world.

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data. We implemented the
framework using existing OHDSI methods for use in the OMOP-CDM, including the
patient-level prediction framework [@Reps2018] and the population-level effect
estimation framework [@Ryan2013] based on new-user cohort design. As a
proof-of-concept we analyzed heterogeneity of the effects of first-line
hypertension treatment: we compared the effect of angiotensin converting enzyme
(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

# Methods
The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A simple overview of the procedure
can be seen in Figure \@ref(fig:figure1).

```{r figure1, echo=FALSE, fig.cap="Caption for figure 1", out.width = '100%'}
# knitr::include_graphics(here("figures/Figure_1.tif"))
grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
```

## Step 1: General definition of the problem 
The typical research aim is: "to compare the effect of treatment $T$ to a
comparator treatment $C$ in patients with disease $D$ with respect to outcomes
$O_1,\dots,O_n$". At least three cohorts are defined:

* A single treatment cohort ($T$) which includes patients with disease $D$
receiving the target treatment of interest. For example, a set of hypertension
patients within a database that receive angiotensin-converting enzyme
inhibitors, followed from the time of initiation until the time of censoring.
* A single comparator cohort ($C$) which includes patients with disease $D$
receiving the comparator (control) treatment. For example, a set of patients
in a database that receive beta blockers, followed from the time of initiation
until the time of censoring.
* One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
developing the outcomes of interest. For example, the
set of patients in a database that have at least one occurrence of acute
myocardial infarction in their record.

## Step 2: Identification of the database 
The aim of this step is the inclusion of databases that represent the patient
population of interest. The inclusion of multiple databases potentially
increases the generalizability of results. Furthermore, the cohorts should
preferably have adequate sample size to ensure precise effect estimation, even
within smaller risk strata.

## Step 3: Prediction 
We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
definition of two essential cohorts: a target cohort and an outcome cohort.

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort . However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
@vanKlaveren2019]. To do this, we developed the patient-level prediction model
in the propensity score-matched subset of the population (1:1), where treatment
assignment is well-balanced. The propensity scores are based on LASSO logistic
regression for modeling the association between treatment assignment and all
available demographics, drug exposures, diagnoses, measurements and medical
procedures. Finally, we need to define the time horizon for which we aim to make
predictions and we need to select the machine-learning algorithm we want to use
to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

## Step 4: Estimation 
We use the patient-level prediction model to divide the target population into
a set of equally-sized risk strata, typically 4 risk quarters. Then, we
estimate propensity scores within risk strata. These propensity scores are used
when estimating treatment effects, either by matching of patients from
different treatment cohorts, by stratification of patients into groups with
similar propensity scores, or by weighing patients’ contribution to the
estimation process. Within risk strata we estimate treatment effect both on the
relative and the absolute scale. It is important to evaluate treatment effects
in both scales, as effect cannot remain constant on both the relative and the
absolute scale at the same time, assuming a non-zero treatment
effect—treatment. Any appropriate method for the evaluation of relative and
absolute treatment effects can be considered, as long as the this is done
consistently in all risk strata.

## Step 5: Result presentation and evaluation 
Our framework provides standardized output for each step of the analysis. The
number of patients and person years by treatment arm along with the number of
outcomes. A performance overview of the derived prediction models, including
discrimination and calibration both in the propensity score matched subset, the
entire population and separately for treated and comparator patients. This is
rather relevant as the performance of the prediction models is directly related
to our ability to single out patient subgroups where treatment may be highly
beneficial or unsafe.  Kent et al [@Kent2016] demonstrated that the event rate and the
discriminative ability of the prediction model can predict very well the
distribution of predicted risk. The lower the event rate and the higher the
c-statistic (given good calibration) result in high risk heterogeneity, thus
making estimated average treatment effects uninformative. In this case, risk
stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

Propensity score distributions by treatment group and covariate balance plots
for each risk stratum. Event rates, hazard ratios and absolute risk differences
in risk strata for a selected outcome, both in tables and in graphs. Hazard
ratios and absolute risk differences for all analyzed outcomes by risk stratum.
Finally, shiny application can be generated to enable easy sharing of the
results.

# Results
As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with a well-established
effectiveness. Beta blockers, even though initially widely used for the
treatment of hypertension, more recent trials and meta-analyses have cast doubt
on their relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines do
not consider them for initial treatment for hypertension while in the EU
guidelines combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line candidate for the treatment [@Khan2006;
@Rabi2020].

## Step 1: General definition of the problem 
We demonstrate the framework with the following research aim: "to compare the
effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients
with established hypertension ($D$) with respect to 9 outcomes
($O_1,\dots,O_9$)". The cohorts are:

* Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
with at least one year of follow-up before treatment start and a recorded
hypertension diagnosis within that year.
* Comparator cohort: Patients
receiving any drug within the beta blocker class with at least one year of
follow-up before treatment start and a recorded hypertension diagnosis within
that year.
* Outcome cohorts: We consider 3 main and 6 safety outcome cohorts.
These are patients in the database with a diagnosis of: acute myocardial
infarction (MI); hospitalization with heart failure; ischemic or hemorrhagic
stroke (efficacy outcomes); hypokalemia; hyperkalemia; hypotension;
angioedema; cough; abnormal weight gain (safety outcomes).

@@

## Step 2: Identification of the databases 
We used the following databases:

* IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
services of retirees (aged 65 or older) in the United States with primary or
Medicare supplemental coverage through privately insured fee-for-service,
point-of-service or capitated health plans. These data include adjudicated
health insurance claims (e.g. inpatient, outpatient and outpatient pharmacy).
Additionally, it captures laboratory tests for a subset of the covered lives.
* IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
Medicaid enrollees from multiple states. It includes hospital discharge
diagnoses, outpatient diagnoses and procedures and outpatient pharmacy claims
as well as ethnicity and Medicare eligibility.  IBM MarketScan Commercial
* Claims and Encounters (CCAE): Data from individuals enrolled in US
employer-sponsored   insurance   health   plans. The data includes adjudicated
health insurance   claims (e.g. inpatient, outpatient, and outpatient
pharmacy) as well as enrollment data from large employers and health plans who
provide private healthcare coverage to employees, their spouses and
dependents. Additionally, it captures laboratory tests for a subset of the
covered lives.

## Step 3: Prediction 
To obtain a target cohort for developing patient-level
predictions we first merged the ACE-inhibitors cohort with the beta blockers
cohort. We then matched patients in the ACE-inhibitor cohort to patients in the
beta blockers cohort on the propensity score. We focused on the efficacy
outcomes (acute MI, hospitalization with heart failure and hemorrhagic or
ischemic stroke) for risk stratification of the patient population. In each
database, for each main outcome, we developed a prediction model. We chose a
time horizon of 2 years after inclusion into the target cohort. We developed the
prediction models using LASSO logistic regression with 3-fold cross validation
for hyper-parameter selection.

## Step 4: Estimation 
We used patient-level predictions to stratify the patient population into 4
risk quarters. We estimated relative and absolute treatment effects for all 9
outcomes of interest. In order to estimate risk quarter-specific treatment
effects we first estimated propensity scores within risk quarters. We then used
the propensity scores to stratify patients into 5 strata.

## Step 5: Result presentation and evaluation 
In the main manuscript we present the results of the analysis in the CCAE
database with stratification based on risk predictions of acute MI. Results of
analyses in the other databases and with other risk stratifications are
included in the supplementary material.

For each outcome and in each risk stratum there were adequate numbers of
patients (Table \@ref(tab:table)). The discriminative ability of the prediction
models was moderate in the matched development subset (c-index 0.76 for acute MI
; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
hospitalization with heart failure was 0.79 and for stroke it was 0.75).
\captionsetup{width=\textwidth}
```{r table, echo=FALSE, results='asis', booktabs=TRUE}
table <- tibble(
  outcome = c(
    rep("Acute myocardial infarction", 4),
    rep("Heart failure (hosp)", 4),
    rep("Stroke (ischemic or hemorrhagic)", 4)
  ),
  riskQuarter = rep(1:4, 3),
  patientsAce = c(
    161099, 204882, 214413, 204167,
    146259, 188006, 218052, 230226,
    146069, 187524, 217070, 226128
  ),
  personYearsAce = c(
    276171, 372197, 393583, 351727,
    249809, 341014, 399394, 400330,
    294484, 340234, 397830, 393861
  ),
  eventsAce = c(
    203, 534, 117, 2095,
    228, 457, 826, 2012,
    299, 554, 947, 1718
  ),
  patientsBeta = c(
    133977, 90193, 80662, 90908,
    126387, 84280, 83421, 98380,
    126264, 84000, 83038, 97628
  ),
  personYearsBeta = c(
    220633, 169231, 150035, 154419,
    206706, 158425, 155222, 169139,
    206453, 157913, 154587, 167810
  ),
  eventsBeta = c(
    135, 321, 535, 1520,
    378, 340, 570, 1773,
    320, 351, 521, 1077
  )
)

@@

Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
ratios decreased) with increasing acute MI risk, resulting in more pronounced
increases of absolute treatment effects (ARD) with increasing acute MI risk
(Figure \@ref(fig:figure2)).

```{r figure2, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment within strata of predicted risk of acute MI. The top panel contains the observed acute rates of ACE-inhibitors and beta blockers within each quarter of predicted acute MI risk. These are derived using the KaplanMeier estimates at 730 days after inclusion. The middle panel, contains the hazard ratios of comparing ACE-inhibitors to beta blockers with regard to acute MI. These are estimated using Cox proportional hazards regression within quarters of predicted acute MI risk. The bottom panel contains absolute risk reduction for ACE-inhibitors compared to beta blockers. These are derived as the difference in Kaplan-Meier estimates at 730 after inclusion. Hazard ratios in the middle panel show a decreasing trend with increasing acute MI risk. Given the rather good discrimination of the prediction model (AUC=0.74), this results in an increasing trend for absolute benefit in favor of ACE-inhibitors with increasing risk.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_2.tif")))
```
Patients in the low risk quarter did not receive absolute treatment benefit (ARD
-0.03%) while absolute risk was 0.54% lower (95% confidence
interval 0.36%—0.71%) for patients in the high-risk quarter. In contrast, the
absolute and relative effects of ACE-inhibitors on safety outcomes (e.g. cough
and angioedema) are approximately constant or even slightly decreasing with
increasing acute MI risk (Figure \@ref(fig:figure3) and \@ref(fig:figure4)).
Similar results were observed in the other two databases (see supplementary
material).

```{r figure3, echo=FALSE, fig.cap="Hazard ratios (relative treatment effects) for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_3.tif")))
```
```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_4.tif")))
```

>>>>

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(here)
library(kableExtra)
d <- function(x, decimals = 2) {
  sprintf(paste0("%1.", decimals, "f"), x) 
}
knit_hooks$set(
  inline = function(x) {
    prettyNum(
      x,
      big.mark = ",",
      decimal.mark = ".",
      preserve.width = "individual"
    )
  }
)
```
\thispagestyle{empty}
\vspace{8mm}

@@

\vspace{10mm}
**Corresponing author**
\singlespacing 
Alexandros Rekkas, MSc

Department of Medical Informatics

Erasmus University Medical Center

3000 CA Rotterdam, P.O. Box 2040

email: a.rekkas@erasmusmc.nl
\onehalfspacing

\vspace{10mm}
**Funding**

This work has been performed in the European Health Data and
Evidence Network (EHDEN) project. This project has received funding from the
Innovative Medicines Initiative 2 Joint Undertaking (JU) under grant agreement
No 806968. The JU receives support from the European Union’s Horizon 2020
research and innovation programme and EFPIA.

\newpage
\newpage
# Abstract {-}
\singlespacing 
One of the aims of the Observation Health Data Sciences and Informatics (OHDSI)
initiative is population-level treatment effect estimation in large
observational databases. Since treatment effects are well-known to vary across
groups of patients with different baseline risk, we aimed to extend the OHDSI
methods library with a framework for risk-based assessment of treatment effect
heterogeneity. The proposed framework consists of five steps: 1) definition of
the problem, i.e. the population, the treatment, the comparator and the
outcome(s) of interest; 2) identification of relevant databases; 3) development
of a prediction model for the outcome(s) of interest; 4) estimation of
propensity scores within strata of predicted risk and estimation of relative and
absolute treatment effect within strata of predicted risk; 5) evaluation and
presentation of results. We demonstrate our framework by evaluating
heterogeneity of the effect of angiotensin-converting enzyme (ACE) inhibitors
versus beta blockers on a set of 9 outcomes of interest across three
observational databases. With increasing risk of acute myocardial infarction we
observed increasing absolute benefits, i.e. from -0.03% to 0.54% in the lowest
to highest risk groups. Cough-related absolute harms decreased from 4.1%
to 2.6%. he proposed framework may be useful for the evaluation of heterogeneity
of treatment effect on observational data that are mapped to the OMOP Common
Data Model. The proof of concept study demonstrates its feasibility in large
observational data. Further insights may arise by application to safety and
effectiveness questions across the global data network.

\vspace{10mm}

**Keywords**: observational data, heterogeneity of treatment effect, risk stratification, subgroup analysis
\newpage 
\doublespacing 
\linenumbers

# Introduction

Interest in understanding how a treatment’s effect varies across patients—a
concept described as heterogeneity of treatment effects (HTE)—has been growing.
This concept is central to the agenda for both personalized (or precision)
medicine and comparative effectiveness research. More formally, HTE has been
defined as non-random variability in the direction or magnitude of a treatment
effect, in which the effect is measured using clinical outcomes
[@KRAVITZ2004]. Usually, analyses focus on the relative scale, where treatment
effects are assessed one at a time in patient subgroups defined from single
covariates, an approach that suffers from low power and multiplicity issues
[@Varadhan2013]. However, even with well-established constant relative
effects, treatment benefit (or harm) may vary substantially on the absolute
scale.

@@

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
common data structure enables analyses at a very large scale. For example, in a
recent study, a large set of first-line treatments for hypertension was compared
with respect to 55 outcomes in a network of databases, including 4.9 million
patients from around the world [@Suchard2019].

We aimed to develop a framework for risk-based assessment of treatment effect
heterogeneity in high-dimensional observational data. We implemented the
framework using existing OHDSI methods for use in the OMOP-CDM, including the
patient-level prediction framework and the population-level effect estimation
framework based on new-user cohort design [@Reps2018; @Ryan2013] . As a
proof-of-concept we analyzed heterogeneity of the effects of first-line
hypertension treatment: we compared the effect of angiotensin converting enzyme
(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

# Materials and Methods

The proposed framework defines 5 distinct steps that enable a standardized
approach for risk-based assessment of treatment effect heterogeneity for
databases mapped to the OMOP-CDM. These are: 1) general definition of the
research aim; 2) identification of the database within which the analyses will
be performed; 3) a prediction step where internal or external prediction models
are used to assign patient-level risk predictions; 4) an estimation step where
absolute and relative treatment effects are estimated within risk strata; 5)
presentation and evaluation of the results. A simple overview of the procedure
can be seen in Figure \@ref(fig:figure1).

```{r figure1, echo=FALSE, fig.cap="(A) Starting from a treatment (top), a comparator (bottom) and an outcome (middle) cohort we estimate the propensity scores on the entire target population. (B) We match patients on the propensity scores and estimate the prediction model. Since we match patients we develop the prediction model on smaller subset of the initial population and, therefore, the number of patients is smaller in B compared to A. (C) We apply the prediction model on the entire population (green: lower 25\\% of the risk distribution; yellow: patients with risk between 25\\% and 50\\% of the risk distribution; orange: patients with risk between 50\\% and 75\\% of the risk distribution; red: patients at risk higher than 75\\% of the risk distribution). (D) We separate in risk subgroups, here quarters. Within risk quarters propensity scores are estimated again and relative and absolute treatment effects are estimated.", out.width = '100%'}
# knitr::include_graphics(here("figures/Figure_1.tif"))
grid::grid.raster(tiff::readTIFF(here("figures/Figure_1.tif")))
```

## Step 1: General definition of the problem

The typical research aim is: "to compare the effect of treatment $T$ to a
comparator treatment $C$ in patients with disease $D$ with respect to outcomes
$O_1,\dots,O_n$". At least three cohorts are defined:

  - A single treatment cohort ($T$) which includes patients with disease $D$
    receiving the target treatment of interest. For example, a set of
    hypertension patients within a database that receive angiotensin-converting
    enzyme inhibitors, followed from the time of initiation until the time of
    censoring.
  - A single comparator cohort ($C$) which includes patients with disease $D$
    receiving the comparator (control) treatment. For example, a set of patients
    in a database that receive beta blockers, followed from the time of
    initiation until the time of censoring.
  - One or more outcome cohorts ($O_1,\dots,O_n$) that contain patients
    developing the outcomes of interest. For example, the set of patients in a
    database that have at least one occurrence of acute myocardial infarction in
    their record.

## Step 2: Identification of the database

The aim of this step is the inclusion of databases that represent the patient
population of interest. The inclusion of multiple databases potentially
increases the generalizability of results. Furthermore, the cohorts should
preferably have adequate sample size to ensure precise effect estimation, even
within smaller risk strata.

## Step 3: Prediction

We adopt the standardized framework for the generation of patient-level
prediction models using observational data that ensures adherence to existing
guidelines [@Collins2015; @Moons2015]. This prediction framework requires the
definition of two essential cohorts: a target cohort and an outcome cohort.

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. However, for risk-based analysis of treatment effects it is
necessary to avoid deferentially fitting the prediction model to patients across
the treatment arm in order to avoid inducing spurious interactions [@Burke2014;
@vanKlaveren2019]. To do this, we develop the patient-level prediction model in
the propensity score-matched subset of the population (1:1), where treatment
assignment is well-balanced. The propensity scores are based on LASSO logistic
regression for modeling the association between treatment assignment and all
available demographics, drug exposures, diagnoses, measurements and medical
procedures. Finally, we need to define the time horizon for which we aim to make
predictions and we need to select the machine-learning algorithm we want to use
to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

## Step 4: Estimation

We use the patient-level prediction model to divide the target population into a
set of equally-sized risk strata, typically 4 risk quarters. Then, we estimate
propensity scores within risk strata. These propensity scores are used when
estimating treatment effects, either by matching of patients from different
treatment cohorts, by stratification of patients into groups with similar
propensity scores, or by weighing patients’ contribution to the estimation
process. Within risk strata we estimate treatment effect both on the relative
and the absolute scale. It is important to evaluate treatment effects in both
scales, as effect cannot remain constant on both the relative and the absolute
scale at the same time, assuming a non-zero treatment effect. Any appropriate
method for the evaluation of relative and absolute treatment effects can be
considered, as long as the this is done consistently in all risk strata.

## Step 5: Result presentation and evaluation

The number of patients and person years by treatment arm along with the number
of outcomes should be reported. A performance overview of the derived prediction
models, including discrimination and calibration both in the propensity score
matched subset, the entire population and separately for treated and comparator
patients should also be reported. This is rather relevant as the performance of
the prediction models is directly related to our ability to single out patient
subgroups where treatment may be highly beneficial or unsafe. Kent et al
[@Kent2016] demonstrated that the event rate and the discriminative ability of
the prediction model can predict very well the distribution of predicted risk.
The lower the event rate and the higher the c-statistic (given good calibration)
result in high risk heterogeneity, thus making estimated average treatment
effects uninformative. In this case, risk stratified analysis of HTE can be more
effective in singling out patient subgroups that stand to benefit (or be harmed)
most by treatment in question.

In every risk stratum propensity score distributions by treatment group and
covariate balance plots should be made available along with event rates, hazard
ratios and absolute risk differences. Finally, the results should be generated
with a standardized format to allow for combination of results from multiple
data sources. The standardized result format also enables the development of
scalable demonstration tools such as R-shiny applications, interactive tables
and figures.

# Results

As a proof of concept, we focus on the comparison of angiotensin converting
enzyme (ACE) inhibitors to beta blockers. ACE inhibitors are among the most
common treatment classes for hypertension, with well-established effectiveness.
Beta blockers, even though initially widely used for the treatment of
hypertension, more recent trials and meta-analyses have cast doubt on their
relative effectiveness [@wiysonge2017beta]. As a result, newer US guidelines
do not consider them for initial treatment for hypertension while in the EU
guidelines combination with other antihypertensive treatments is recommended
[@Whelton2018; @Williams2018]. However, another meta-analysis suggested that
the efficacy profile of beta blockers is similar to other major treatment
classes in younger hypertensive patients and, thus, countries like Canada still
include them as a first-line treatment candidate [@Khan2006; @Rabi2020].

## Step 1: General definition of the problem

We demonstrate the framework with the following research aim: "to compare the
effect of ACE-inhibitors ($T$) to the effect of beta blockers ($C$) in patients
with established hypertension ($D$) with respect to 9 outcomes
($O_1,\dots,O_9$)". The cohorts are:

  - Treatment cohort: Patients receiving any drug within the ACE-inhibitor class
    with at least one year of follow-up before treatment initiation and a
    recorded hypertension diagnosis within that year.
  - Comparator cohort: Patients receiving any drug within the beta blocker class
    with at least one year of follow-up before treatment initiation and a
    recorded hypertension diagnosis within that year.
  - Outcome cohorts: We consider 3 main and 6 safety outcome cohorts. These are
    patients in the database with a diagnosis of: acute myocardial infarction
    (MI); hospitalization with heart failure; ischemic or hemorrhagic stroke
    (efficacy outcomes); hypokalemia; hyperkalemia; hypotension; angioedema;
    cough; abnormal weight gain (safety outcomes).

@@

## Step 2: Identification of the databases

We used the following databases:

  - IBM MarketScan Medicare Supplemental Beneficiaries (MDCR): Represents health
    services of retirees (aged 65 or older) in the United States with primary or
    Medicare supplemental coverage through privately insured fee-for-service,
    point-of-service or capitated health plans. These data include adjudicated
    health insurance claims (e.g. inpatient, outpatient and outpatient
    pharmacy). Additionally, it captures laboratory tests for a subset of the
    covered lives.
  - IBM MarketScan Medicaid (MDCD): Adjudicated US health insurance claims for
    Medicaid enrollees from multiple states. It includes hospital discharge
    diagnoses, outpatient diagnoses and procedures and outpatient pharmacy
    claims as well as ethnicity and Medicare eligibility.
  - IBM MarketScan Commercial Claims and Encounters (CCAE): Data from
    individuals enrolled in US employer-sponsored insurance health plans. The
    data includes adjudicated health insurance claims (e.g. inpatient,
    outpatient, and outpatient pharmacy) as well as enrollment data from large
    employers and health plans who provide private healthcare coverage to
    employees, their spouses and dependents. Additionally, it captures
    laboratory tests for a subset of the covered lives.

## Step 3: Prediction

To obtain a target cohort for developing patient-level predictions we first
merged the ACE-inhibitors cohort with the beta blockers cohort. We then matched
patients in the ACE-inhibitor cohort to patients in the beta blockers cohort on
the propensity score. We focused on the efficacy outcomes (acute MI,
hospitalization with heart failure and hemorrhagic or ischemic stroke) for risk
stratification of the patient population. In each database, for each main
outcome, we developed a different prediction model. We chose a time horizon of 2
years after inclusion into the target cohort. We developed the prediction models
using LASSO logistic regression with 3-fold cross validation for hyper-parameter
selection.

## Step 4: Estimation

We used patient-level predictions to stratify the patient population into 4 risk
quarters. We estimated relative and absolute treatment effects for all 9
outcomes of interest. In order to estimate risk quarter-specific treatment
effects we first estimated propensity scores within risk quarters. We then used
the propensity scores to stratify patients into 5 strata.

## Step 5: Result presentation and evaluation

Here we present the results of the analysis in the CCAE database with
stratification based on risk predictions of acute MI. Results of analyses in the
other databases and with other risk stratifications are included in the
supplementary material.

For each outcome and in each risk stratum there were adequate numbers of
patients (Table \@ref(tab:table)). The discriminative ability of the prediction
models was moderate in the matched development subset (c-index 0.76 for acute MI
; 0.79 for hospitalization with heart failure; 0.74 for stroke; ), in the
general population (c-index 0.74 for acute MI; 0.77 for hospitalization with
heart failure; 0.73 for stroke), in the treatment cohort (c-index for acute MI
it was 0.71, for hospitalization with heart failure was 0.76 and for stroke it
was 0.72) and in the comparator cohort (c-index for acute MI it was 0.79 for
hospitalization with heart failure was 0.79 and for stroke it was 0.75).
\captionsetup{width=\textwidth}
```{r table, echo=FALSE, results='asis', booktabs=TRUE}
table <- tibble(
  outcome = c(
    "Acute myocardial infarction",
    rep("", 3),
    "Heart failure (hosp)",
    rep("", 3),
    "Stroke (ischemic or hemorrhagic)",
    rep("", 3)
  ),
  riskQuarter = rep(1:4, 3),
  patientsAce = c(
    161099, 204882, 214413, 204167,
    146259, 188006, 218052, 230226,
    146069, 187524, 217070, 226128
  ),
  personYearsAce = c(
    276171, 372197, 393583, 351727,
    249809, 341014, 399394, 400330,
    294484, 340234, 397830, 393861
  ),
  eventsAce = c(
    203, 534, 117, 2095,
    228, 457, 826, 2012,
    299, 554, 947, 1718
  ),
  patientsBeta = c(
    133977, 90193, 80662, 90908,
    126387, 84280, 83421, 98380,
    126264, 84000, 83038, 97628
  ),
  personYearsBeta = c(
    220633, 169231, 150035, 154419,
    206706, 158425, 155222, 169139,
    206453, 157913, 154587, 167810
  ),
  eventsBeta = c(
    135, 321, 535, 1520,
    378, 340, 570, 1773,
    320, 351, 521, 1077
  )
)

@@

Relative treatment effects of ACE-inhibitors vs beta blockers increased (hazard
ratios decreased) with increasing acute MI risk, resulting in more pronounced
absolute risk differences (ARD) with increasing acute MI risk (Figure
\@ref(fig:figure2)).

```{r figure2, echo=FALSE, fig.cap="Overview of heterogeneity of ACE-inhibitors treatment within strata of predicted risk of acute MI. The top panel contains the observed acute rates of ACE-inhibitors and beta blockers within each quarter of predicted acute MI risk. These are derived using the KaplanMeier estimates at 730 days after inclusion. The middle panel, contains the hazard ratios of comparing ACE-inhibitors to beta blockers with regard to acute MI. These are estimated using Cox proportional hazards regression within quarters of predicted acute MI risk. The bottom panel contains absolute risk reduction for ACE-inhibitors compared to beta blockers. These are derived as the difference in Kaplan-Meier estimates at 730 after inclusion. Hazard ratios in the middle panel show a decreasing trend with increasing acute MI risk. Given the rather good discrimination of the prediction model (AUC=0.74), this results in an increasing trend for absolute benefit in favor of ACE-inhibitors with increasing risk.", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_2.tif")))
```

Patients in the low risk quarter did not receive absolute benefit (-0.03%)
while absolute risk was 0.54% lower (95% confidence interval 0.36%—0.71%) for
patients in the high-risk quarter. In contrast, the absolute and relative
effects of ACE-inhibitors on safety outcomes (e.g. cough and angioedema) are
approximately constant or even slightly decreasing with increasing acute MI risk
(Figure \@ref(fig:figure3) and \@ref(fig:figure4)). Similar results were observed
in the other two databases (see supplementary material).

```{r figure3, echo=FALSE, fig.cap="Hazard ratios (relative treatment effects) for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_3.tif")))
```

```{r figure4, echo=FALSE, fig.cap="Absolute risk reduction for the main and safety outcomes, estimated by fitting stratified Cox regression models within quarters of predicted risk of acute myocardial infarction (MI). The four risk quarters (Q1-Q4) are defined using the internally developed model for acute MI", out.width = '100%'}
grid::grid.raster(tiff::readTIFF(here("figures/Figure_4.tif")))
```
